\documentclass[12pt]{article}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}
\oddsidemargin=0.0in
\evensidemargin=0.0in
\textwidth=6.3in
\headheight=0.0in
\topmargin=0.0in
\textheight=8.9in
\title{Minimising Portfolio Loss in Portfolio Construction}
%\date{16-12-2021}
\author{Colin M. Smith BITA Risk}
\begin{document}
\maketitle
\tableofcontents
\pagebreak
\doublespacing
\section{Introduction}
This document attempts to show how to use portfolio loss minimisation in the
portfolio enhancement process. We start by defining $Loss$ and compare how
the predicted changes to a portfolio, that have been achieved by adopting a minimal $Loss$ as opposed to a minimal $Risk$ 
strategy, are different. 
We indicate how the minimisation of
$Loss$ could replace or augment $Risk$ minimisation in the construction steps but also admit
that this is likely to need more execution time.
Finally a more tractable $`$Gain Loss with a twist' is described. We propose using
the factor model to find predicted asset return series and use these to define $predicted\ Loss$ and $Gain$. 
By minimising portfolio predicted $Loss$ in the construction, we should get a faster calculation since 
the predicted power of the risk model can be delivered by as few as three months of factor premia returns' data.
This should control dispersion in a different way to modelled risk as it does not penalise predicted gains.

\section{Definition of Portfolio Gain and Loss}
Over the history of a portfolio it is usual to monitor return each period and the returns of the underlying assets.
From one period to the next, the portfolio's return will change and to characterise this change we can define its gain or loss.
To make the gains and losses a little more configurable we define 
gain $G$ and loss $L$ with respect to a target return $R$ which could be positive or negative.
Suppose the portfolio return series at time $t$ is $r(t)$ then
\begin{eqnarray}
    G = \sum_t {\textbf{ max} }(r(t) - R,0)
\end{eqnarray}
\begin{eqnarray}
    L = \sum_t {\textbf{ max} }(R-r(t),0)
\end{eqnarray}
This means that in period $t$, if $R-r(t) > 0$ there is a loss, and if $R-r(t) < 0$ there is a gain.
If we define a loss period as one for which $R-r(t) > 0$,
\begin{eqnarray}
  {\rm Total\ Loss},\ L= \sum_{t\ \in\ \rm {loss\atop periods}} R-r(t)
\end{eqnarray}
$G$ and $L$ are both non-negative and
\begin{eqnarray}
 {\rm   total\ portfolio\ return}= G - L + RT
\end{eqnarray}
Thus expected return, gain and loss are not independent. An optimisation to maximise the utility $G- \lambda L$
would be set up as a miminisation of utility $-\frac{\gamma}{1-\gamma}\sum\alpha_i w_i + L$ (which is 
similar in spirit to the mean variance utility $-\frac{\gamma}{1-\gamma}\sum\alpha_i w_i + 0.5 \sum w_i C_{ij}w_j$).
$\gamma$ is the usual return aversion multiplier we use and can take a value in the interval $[0,1)$.

If there are $n$ assets in the portfolio and each asset $i$
has a weight $w_i$ and a return series $s_i (t)$ then
\begin{eqnarray}
    r(t) = \sum_{i=1}^{n} w_i s_i(t)
\end{eqnarray}
The expected return for asset $i$, $\alpha_i$ may be calculated from
mean\footnote{We calculate $\alpha$ using arithmetic mean. $\alpha_i$ and target rate are values per-period.} 
return of $s_i(t)$.
\begin{eqnarray}
    \alpha_i =  \frac{1}{T}\sum_{t=1}^{T} s_i(t)
\end{eqnarray}
where $T$ is the number of periods.

\section{Comparison of Optimal Portfolio Weights in Risk and Loss Minimisation}

Following standard portfolio construction methods, the practioner seeks to find a portfolio that 
will give a certain amount of return for smallest risk. Risk is not the only way to measure 
dispersion though. Instead of minimising risk, we could find the 
lowest possible loss to attack dispersion. For illustration purposes, consider a small portfolio of nine assets
for which we have a return series for each asset.
The following bar chart shows the outcome weights for two portfolio optimisations. There are $423$ data periods in the
return series. 


\begin{tikzpicture}
    \begin{axis}
        [title={Min Loss weights compared with Min Risk weights},
        ybar,height=2in,width=6.3in, symbolic x coords={1,2,3,4,5,6,7,8,9},
        legend style={at={(0.5,-0.2)},        anchor=north,legend columns=-1}]
        \addplot coordinates {(1,0)
        (2,0) 
        (3,0)
        (4,0.3854726513396513)
        (5,0)
        (6,0.3950537500804109)
        (7,0)
        (8,0)
        (9,0.21947359857993776)};
        \addplot coordinates {
            (1,0.2131871336385196)
            (2,0)
            (3,0.008529985052857314)
            (4,0.07179520200177866)
            (5,0.13819750244263093)
            (6,0.046937456187809756)
            (7,0.0047705423361060345)
            (8,0.07143877895977421)
            (9,0.4451433993805218)};
            \legend{Min Risk weights,Min Loss weights}
    \end{axis}
\end{tikzpicture}


\begin{tikzpicture}
    \begin{axis}[title={Asset alpha values},
        ymin=-0.002,
        ymax=0.01,
        scaled y ticks=false,yticklabel style={/pgf/number format/1000 sep=},
        ybar,height=2in,width=6in,symbolic x coords={1,2,3,4,5,6,7,8,9}]
        \addplot coordinates{
            (1,0.0011707063636410724)
            (2,-0.0018399458749987034)
            (3,-0.001547403488527072)
            (4,0.003674758692542117)
            (5,-0.00016312751189371318)
            (6,0.003627629196094118)
            (7,-0.001805179090833558)
            (8,-0.000010932381824192851)
            (9,0.009797864161556428)}[color=green];
    \end{axis}
\end{tikzpicture}


In both optimisations\footnote{The weights are constrained to satisfy the budget $\sum w_i = 1$} the return is 
constrained to $0.005$; the blue bars show the weights
for mininum risk ($0.058$) which gives a portfolio with total loss of $1.34$ and the pink bars show
the weights for minimum total loss ($1.15$) with a risk of $0.085$. The loss is defined with respect to a target rate per period of 
$0.0004$. The risk quoted is historic risk and is 
calculated directly as the standard deviation of the portfolio return series. We don't annualise risk and return or 
express them as percentage values here.
Notice that the $Loss$ minimising portfolio achieves its constrained return by putting most weight in the
asset with highest alpha, and uses a diverse selection of the other assets to make up the remaining 
return. It even chooses assets with negative alpha\footnote{This is a consequence of the budget constraint}. 
The $Risk$ minimising portfolio seems less diverse and makes less use of the high alpha asset.
From experience, it seems this happens when the utility function contains just historic return and variance or loss terms,
subject to linear constraints. If the linear part of the utility function is non-zero but adopts expected alphas which differ from historic mean 
enough then the differences between loss and risk minimisation are not so great.
By definition, the minimum loss 
portfolio must have more risk than the minimum risk portfolio, but we argue that this extra risk is allowable 
because the loss is minimal.

If we just maximised return subject to the budget, the portfolio would have all weight in the last 
asset and the return would be $0.0098$.

\section{Minimising Loss in the Enhancements}
The portfolio enhancement process consists of a host of regulatory constraints and other constraints 
designed to improve portfolio expected return. It is necessary to try to control the uncertainty in the return usually 
by managing the risk.
A straightforward way to adopt loss optimisation might be to replace the risk optimisation with loss minimisation.
Over the entire history the weights could be chosen to minimise total portfolio $Loss$ subject to the constraints.
If we want to incoporate gain maximisation as well we add historic asset means to the utility function.
We require full asset return history to define $Loss$ at each optimisation step,
and this could 
result in a slow optimisation process due to the vast amount of history\footnote{It would be possible 
to use a risk constraint as well but this would mean even longer execution time.}.  An advantage of
minimising $Loss$ over minimising $Risk$ is that the resulting weights could
be more diversifying. Looking at the simple graph for 9 stocks in the previous section we see that
$Loss$ minimisation results in higher weight in the high alpha asset with many 
smaller changes in the other assets to get the lowest loss. ($Risk$ optimisation gives a less diverse
result and a lower weight in the high alpha asset). We conjecture from this that $Loss$ minimisation 
would make less impact on the strength of the factor premia terms, than $Risk$ minimisation although by definition it cannot
manage risk as well. This may not matter though; the difference in weight outcomes
highlights that the gains and losses in the optimisation are not symmetrically
distributed. ($Risk$ minimisation penalises gains and losses from the mean in the same way.) If they were the same, minimising $Loss$ would be very similar to minimising historic $Risk$, (although
we do impose our own asymmetricity by having a non-zero target rate.) 

It is important
to note that the factor premia must not be constrained too high, otherwise there is little scope 
for minimising $Loss$ (or $Risk$ either). In the example in the previous section,
if we had constrained the return to be $0.0098$ the only possible outcome has $100\%$ in 
the last asset irrespective of whether the utility contains risk or loss.

The simple example explicitly contains an alpha constraint, but alpha as such does not need to 
be entered into the constraints. We include it in the utility function so as to maximise gain if needed. 
Instead of constraining alpha we could include the momentum 
vector for example in the constraints instead of the alpha vector. We used the alpha constraint in the example as a general proxy for all 
constraints of interest used together.
\section{Minimising Modelled Loss}
The calculation thus far might be thought of as $`$traditional gain loss'; it
maximises $historic\ (gain-\lambda\ loss)$ subject to the portfolio constraints. Usually the risk optimised in 
portfolio enhancement is modelled risk. If we want to adopt loss minimisation, it might be better to use $modelled\ loss$ and $modelled\ gain$ here rather than historic values 
and add $modelled\ asset\ means$
in the utility to optimise gain.

The risk model estimation process uses factor premia returns $p_{i}(t)$ in the regression
with asset returns. If there are $n_f$ factor premia, the modelled asset return $\hat{r}_{k}(t)$ for asset $k$ may be written
\begin{eqnarray}
    \hat{r}_{k}(t) = \sum_{i=1}^{i=n_f}\beta_{ki} p_{i}(t)
\end{eqnarray}
\begin{eqnarray}
    \hat{\alpha}_k = \frac{1}{T}\sum_{t=1}^T \hat{r}_{k}(t)
\end{eqnarray}
at time $t$, where $p_i$ is the i'th factor return and $\beta_{ki}$ is the usual factor loading matrix of regression coefficients.
$\hat{\alpha}_k$ is the modelled expected return for the $k$th asset.
This equation neglects the specific return which is not explained by the factors. The specific returns aren't of interest here because we want 
to maximise the effects due to just the factors. It is only necessary to include the factors which we expect 
to drive high returns.
Due to observed cycle conditions exhibited by the factors, a three month forecast window\footnote{One month is observed to be too short an estimation period.} of factor premia data may be enough
to get a good prediction of future return. Instead of incorporating the full history of
asset returns $r_k(t)$ in the optimisation it may be enough to use the modelled predictions
$\hat{r}_{k}(t)$ over a much reduced time horizon. The shorter the history for defining the
loss, the faster the optimisation, so there is a practical reason for this as well as
a theoretical one. The cyclical events captured by the factors may be correlated negatively or positively with 
the true portfolio return. This is shown by the sign of $\beta$ and it is therefore easy to include all factors together 
which have a significant impact on return.
Judging from the weights obtained in the simple 9 stock loss minimisation, we would expect 
that minimising the loss defined by $\hat{r}_{k}(t)$, subject to factor premia constraints, would favour mainly weights 
with high exposure to the factor premia together with diversifying weights that enable the lowest modelled loss.
\end{document}

